Hamiltonian mechanics is conservation of information entropy

### 2019-01-10 Articles/HamiltonianMechanicsIsConservationOfEntropy/article.tex
Incorporating final comments

\title{Hamiltonian mechanics is conservation of information entropy}
\author{Gabriele Carcassi, Christine A. Aidala}

@@

\date{\today}

@@

The first is that they concentrate on the mathematical structure which, unfortunately, is not enough to characterize a physical theory. They all note how the same physical system can be described by different mathematical frameworks. For example, a classical system can be characterized by position and velocity in the Lagrangian framework and by position and momentum in the Hamiltonian framework. They do not note, though, that the converse is also true: the same mathematical equation can represent two different physical systems. For example, the equation $ma + bv = 0$ in the context of Newtonian mechanics can either represent a body in an inertial frame under linear drag or a body in a non-inertial frame with no forces acting on it. If $t_{in}$ is time in the inertial frame, the time $t$ in the non-inertial frame is such that  $\frac{dt}{dt_{in}} = e^{\frac{b}{m}t}$. Mathematical equivalence does not imply physical equivalence, which is especially problematic for \cite{Barrett2} since it focuses on the former. In other words, we can have the same physics with different math and the same math with different physics. The connection between math and physics is many-to-many because mathematics captures only the syntax of a particular description without its semantics, which at that point cannot be recovered.

As we do not have a one-to-one map between math and physics, fixating on the mathematical structure may even confuse us. For example, it is known\cite{AllHamFreeParticle} that any Hamiltonian system is, at least locally, equivalent to a free particle. That is, we can always find a canonical transformation (i.e. a transformation that does not change the equation of motion) such that any Hamiltonian system, in the new coordinates, has the motion of a free particle for a finite amount of time. As \cite{North} notes, only frame-independent relationships should be taken as fundamental. Does that mean that the Hamiltonian itself is not fundamental because it is not invariant under canonical transformations? Does it mean that all Hamiltonian systems are the same system, since there is always a set of canonical coordinates in which $H=p^2/2m$? It is also known\cite{AllSystemsAreHam} that for any first-order system of ordinary differential equations we can find an appropriate symplectic structure such that the system can be written as a Hamiltonian system. Does that mean that, in the end, all systems are physically equivalent, regardless of what they describe? The point is, we cannot know whether our findings about mathematical structures are physically significant if we do not have a precise link between the mathematical symbols and their physical meaning.

This leads us to the second problem we believe the previous works may have: when trying to attach meaning the thinking and intuition is still Newtonian. This is particularly evident in \cite{Curiel} where physical ideas provide the starting point to get to Lagrangian dynamics. The starting point is that a classical system is one that, roughly, obeys Newton's second law and therefore we have, for a free particle, $\dot{x} = v$ and $\dot{v} = 0$. Unfortunately this is only true in an inertial frame. As \cite{North} notes, both Hamiltonian and Lagrangian mechanics are coordinate independent: the equations are valid in all reference frames even non-inertial ones. In fact, that is probably the most notable advantage of the frameworks, that fictitious forces are automatically handled. But none of the Newtonian concepts are coordinate independent: inertial frames, absence of forces and even conservation of energy are all frame dependent. This, again, leads to confusion and we may miss a key element: in the transition from Newtonian to Lagrangian/Hamiltonian mechanics we expand the reference frames in which the equations are valid (from inertial frames to all frames) at the expense of restricting the class of systems we can describe (e.g. dissipative forces in general require a modification of the Euler-Lagrange equation typically using the Rayleigh dissipation function). The point is, Newtonian concepts may prevent us from fully understanding Hamiltonian and Lagrangian mechanics.

It should be clear that we do not fault the authors. We fault ourselves, the physicists. Even just a century ago, there was a push in our field to find fundamental principles or laws that could serve as a starting point for the different theories. Newtonian mechanics, thermodynamics and special relativity serve as good examples. Nowadays, unfortunately, many of the current fundamental theories, including Lagrangian, Hamiltonian, quantum mechanics and quantum field theory take as their starting point a particular mathematical structure. That is, we do not know what the value of a Lagrangian means or what the symplectic form represents; we just take them as given. Moreover, a lot of work in theoretical physics nowadays starts by exploring novel abstract mathematical structures and then trying to see whether some physical meaning or prediction can be extracted. We are not going to debate here how and why this shift happened. We merely note that it leaves the semantics of our theories ill defined.\footnote{Which is a nice way to say that we don't really know what we are talking about.}

@@

As all these problems stem from a lack of physical principles that can serve as a foundation for classical Hamiltonian and Lagrangian mechanics, we turn to a recently published work\cite{AoPPhy1} that aims to identify such starting points.\footnote{We should also mention \cite{Maudlin} as another attempt to formally derive mathematical structure from physical ideas.} Inspired by those insights, we fully develop another that is only briefly hinted at in that work: the idea that \textbf{a distribution over a manifold conserves information entropy over time if and only if its elements evolve according to Hamiltonian mechanics.} That is, the single requirement of information entropy conservation not only gives us Hamiltonian dynamics but also the structure of phase space (i.e. a symplectic structure where variables are organized in conjugate pairs).

We believe this insight may be of interest to the philosophy of science community for at least two reasons. First because it helps to clearly characterize what Hamiltonian mechanics describes, providing a clear physical meaning to the mathematical structures used. This should be at least of interest to that part of the philosophy community interested in the foundations of classical mechanics. Second because information entropy has a deep connection to the foundations of thermodynamics, therefore the strong link we establish here between information entropy and Hamiltonian mechanics should be of interest to an even wider audience.

@@

Before proceeding, though, let us give a quick introduction so that we have a clear understanding about what information entropy represents. In the physics community, for example, the concept is often vaguely referred to as ``knowledge" (and sometimes as ``lack of knowledge") which is incorrect and misleading.\footnote{We believe this characterization may be due to Jaynes\cite{Jaynes} who first introduced the concept of information theory within thermodynamics. Note that we are only going to talk about information entropy and not thermodynamic entropy. We do believe there is a link between the two, but it is not the one Jaynes gives, which we believe to be unphysical. We leave that discussion for later works.} Information entropy was introduced by Shannon\cite{Shannon} to solve specific engineering problems and therefore, like distance or mass, information entropy is a number that measures a well defined quantity and should be treated as such. What he needed was a way to measure how much information a particular channel is able to transfer and therefore a way to quantify precisely information as number of bits. So the idea is that we have a source that sends messages to a destination, and the message is encoded in zeros and ones. Information entropy is the optimal average number of zeros and ones that must be used to encode the messages. If the source sends fewer bits per message than required by the information entropy, some information will be lost. If the source sends more bits per message than required by the information entropy, some information will be redundant.

@@

In general, if $I$ is the number of questions, the number of bits, then $C = 2 ^ I$ are the number of cases one can distinguish with those questions. On the other hand, suppose for a particular value $\xi_0$ of a discrete variable we have $\rho(\xi_0)=1/5$, then we know that we'd select $\xi_0$ one in 5 cases. That is, the normalized distribution is the inverse of the number of cases: $\rho(\xi_0) = \frac{1}{C}$.  Therefore the information needed to identify a particular case is $I(\xi^a)=\log \frac{1}{\rho(\xi^a)}$. If we take the expectation value of this expression over the distribution and change the base of the logarithm, we have:
\begin{equation}
I(\rho(\xi^a)) = \int \rho(\xi^a) \ln \left(\frac{1}{\rho(\xi^a)}\right) d\xi^1 d\xi^2 ... d\xi^n =-\int \rho(\xi^a) \ln (\rho(\xi^a) d\xi^1 d\xi^2 ... d\xi^n
\end{equation}
which is the information entropy of the distribution. This represents the average number of bits one needs to put into a message to tell someone else the value of the element of the distribution that was picked at random.

@@

The above discussion, while not a formal derivation, should give a sense of what the expression signifies: it measures the \emph{additional} information that would be needed to identify the value associated to an element picked at random from a known population. Distribution that are more spread will have a higher information entropy. What is also important is that it is a precisely defined quantity and that there is only one way to define it. It is the only expression that satisfies three simple requirements (continuity, monotonicity and additivity) that are necessary to measure information, and that is how Shannon originally derived it. Insofar that we want to talk about amounts of information we have only one way to do so.

@@

We now turn our attention to what \cite{AoPPhy1} calls an infinitesimally reducible system. This is a system that can be repeatedly divided into smaller and smaller parts and for which giving a description of the full system is equivalent to giving a full description for all parts. We call particles the limit of this recursive subdivision. Note that particles here are not point-like: they have an infinitesimal size for each dimension along which the system was subdivided. Under this premise each particle is assigned a state from a state space $\mathcal{S}$ and the state of the whole system will be described by a distribution over those particle states. The elements of our distribution are the particles and the property that characterizes each element is its state.

@@

While the physical setting is clear enough, it introduces a puzzle. States will be themselves identified by quantities, which we call state variables. But for states to be physically objective, they must be independent of reference frame. For example, if $s$ is the state, $A$ and $B$ are two reference frames and $\xi^a$ and $\hat{\xi}^b$ are the state variables one would use in the respective reference frames, we must have $s(\xi^a) = s(\hat{\xi}^b)$.  Therefore, since the distribution is defined on the state itself, we must have $\rho(s(\xi^a))=\rho(s(\hat{\xi}^b))$. In this case, it seems we \emph{do} want the density to be defined on the actual space $\mathcal{S}$ and it seems it will have to be invariant under an arbitrary change of state variables. Accordingly, if we want densities and information entropy to be conserved over time, we first need to make sure these quantities are defined in a way that is observer independent, since otherwise they would be conserved in one frame but not in another. But this seems in apparent contradiction with the previous section and we need to clarify that.

@@

If we want the time evolution to transport  densities and to conserve information entropy, the Jacobian must be unitary and therefore $S$ is divergence-free and admits a potential. We have
\begin{equation}
\label{Potential_Hamilton}
\begin{aligned}
S &= \left(\frac{\partial H}{\partial p}, - \frac{\partial H}{\partial q}\right) \\
\frac{dq}{dt} &= \frac{\partial H}{\partial p}  \\
\frac{dp}{dt} &= - \frac{\partial H}{\partial q}  \\
\end{aligned}
\end{equation}
and we recognize Hamilton's equations. Note that the reverse is true as well: any Hamiltonian evolution will transport densities and conserve information entropy due to Liouville's theorem.\footnote{Again, it is not just that the density and the areas are conserved in time: it is that they are the same for all canonical transformations including change of coordinates. That is, we have an objective way to compare them.} Therefore \textbf{Hamiltonian mechanics is precisely the evolution of (classical) distributions over a manifold that preserves information entropy and densities}.

@@

%Now, the fact that we can use a tangent bundle to track our particle states does \emph{not} mean that conjugate momentum is a one form, in the sense of a map from a vector to a number. The state space is just a manifold, a set of points identified by $n$ continuous variable, and the structure it has comes from how units transform, not because the variables identify separate objects. Unfortunately mathematical structures strip out the units, the knowledge of the double role of coordinates is lost and then a different mathematical concept is used, one that has the same mathematical features of our particle state space but an altogether different meaning and construction. Changing state variables from position and momentum to position and velocity is not, as one may mistakenly understand from the math, a change of objects from a cotangent bundle to a tangent bundle: it is relabeling the same states much like changing spatial coordinates relabels positions. This change of variables, though, is not in general a canonical transformation which means the density will not correspond to the invariant one. But it is still a valid change of state variables and in fact, giving that velocity is gauge invariant, a very useful one. It allows to write Poisson brackets and find invariants of motion that are readily physically meaningful\cite{Holten}.

@@

Now, the fact that we can use symplectic manifolds, or tangent bundles, to track our particle states does \emph{not} mean that conjugate momentum is a one form, in the sense of a map from a vector to a number, defined on the coordinate space. The state space comes as one whole manifold, a set of points identified by $n$ continuous variables $\xi^a$, and the structure it has comes from how units transform, not because the variables identify separate objects. Unfortunately, mathematical structures strip out the units, the knowledge of the double role of coordinates is lost and then a different mathematical concept is used, one that has the same mathematical features of our particle state space but an altogether different meaning and construction. In the same vein, changing state variables from position and momentum to position and velocity is not, as one may mistakenly understand from the math, a change of objects from a cotangent bundle to a tangent bundle: it is simply relabeling the same states much like changing spatial coordinates relabels positions. This change of variables, though, is not in general a canonical transformation which means it does not preserve $\omega$ so the density will not correspond to the invariant one. But it is still a valid change of state variables and in fact, given that velocity is gauge invariant, a very useful one. It allows one to write Poisson brackets and find invariants of motion that are readily physically meaningful\cite{Holten}.

@@

The usual approach posits point-like particles, with a particular state space and equations of motion. Then, coincidentally, finds this setting convenient for statistical mechanics because it has nice properties. In retrospect, this should sound backwards. We believe the approach presented here is preferable, first of all, because it justifies with a single insight the state space, the evolution and why that state space is so compatible with statistical mechanics.

Second, it maps better to what we do in classical physics. We study composite systems (planets, cannonballs, beads on a wire, fluids) which we sometimes pretend we can recursively divide. None of those objects, in fact, has a single position or momentum: only the center of mass has a single position and momentum but those are precisely the expectation values $\int q \rho(q,p) dq dp$ and $\int p \rho(q,p) dq dp$ over the distribution. Under this light, asking whether classical mechanics is tenable coincides with asking whether we can really keep dividing our object indefinitely. This approach also fails more gracefully. If we started with the idea of point-like objects and that idealization fails, any aggregate will also fail. But since we start with the macroscopic object and subdivide, the macroscopic object still remains what it is even if the idealization of infinite subdivision fails.

Third, this new approach is conceptually more compatible with the other physical theories. We have seen how elements of special relativity (e.g. coordinate invariance, momentum as covector), statistical mechanics (e.g. joint and marginal distributions), quantum mechanics (e.g. wave number) and classical Hamiltonian mechanics (e.g. phase space, Poisson brackets) already emerged naturally. Furthermore, note that in general relativity point-like particles would be singularities that puncture space-time and would not move along geodesics, while particles as infinitesimal parts would. Note that if the evolution is deterministic and reversible, it means the evolution of the system depends only on itself, which means that the system is isolated. That is, we have an association between isolated system and conservation of energy.\footnote{To be precise, conservation of energy is not an invariant property. The fact that a Hamiltonian exists, though, is an invariant property.} This sounds strikingly close to the first law of thermodynamics.

The larger point is that physics is one and it cannot truly be understood piecemeal. Nature does not separate between dynamical systems and thermodynamics, between topologies and $\sigma$-algebras, between classical systems and quantum systems. The boundaries between all these disciplines are as much unfortunate as they are artificial. There is no real understanding of one without the other.

Having a coherent set of ideas, a coherent picture across the different areas is not only intellectually more satisfying, but it allows us to switch perspective when the intuition from one fails. As a practical example, we can show how understanding the link between Hamiltonian mechanics and information theory leads to a classical analogue for the uncertainty principle.

@@

As we saw, Hamiltonian dynamics not only conserves the energy of individual particles, but it also conserves the information entropy of the distribution as a whole. Intuitively, we can see that there is some link between the spread of a distribution and the number of bits required to identify an element. For example, if we have two uniform distributions with different ranges, then Alice will need to give Bob more information to identify the value from the bigger range. So we ask: what is the distribution that minimizes the spread given a certain amount of information entropy? If $I$ is the set amount of information entropy and $\sigma_q^2 \sigma_p^2 \equiv \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$ is the spread, we can use Lagrange multipliers to find out.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1) \\ &+ \lambda_2(- \int \rho \ln \rho \, dqdp - I)\\ 
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 + \\ &\lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\end{align*}
We solve the multipliers and have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I = &\ln (2\pi\sigma_q\sigma_p) + 1
\end{align*}
The distribution that minimizes the spread, then, is the product of two independent Gaussians. As the entropy is conserved during Hamiltonian evolution the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I - 1) / 2 \pi 
\end{align*}
Intuitively, Hamiltonian mechanics is deterministic and reversible evolution for a distribution and its elements, and therefore it cannot reduce the spread of the distribution indefinitely. If it did, it would start concentrating the density, meaning that multiple states would be mapped to a single state and the dynamics would not be reversible. In this form $I$ is not a fundamental constant but a parameter that depends upon the initial condition. While this is already striking, \cite{AoPPhy1} shows a simple argument to turn this into something more.

Suppose that we have a distribution where the particles move non-derministically. That is, their motion is partly caused by some other agent in a way that can be characterized by random motion. Now suppose our distribution is at an equilibrium with that random motion. Then the particles change state randomly but the overall distribution remains the same. That is, the random motion simply reshuffles the elements. In this case the minimum achievable for the information entropy of the distribution depends on the random motion itself. The stronger it is, the more spread the distribution has to be, the higher the minimum threshold for the information entropy.

>>>>

\title{Hamiltonian mechanics is conservation of information entropy}
%\author{Gabriele Carcassi, Christine A. Aidala}

@@

%\date{\today}

@@

The first is that they concentrate on the mathematical structure which, unfortunately, is not enough to characterize a physical theory. They all note how the same physical system can be described by different mathematical frameworks. For example, a classical system can be characterized by position and velocity in the Lagrangian framework and by position and momentum in the Hamiltonian framework. They do not note, though, that the converse is also true: the same mathematical equation can represent two different physical systems. For example, the equation $ma + bv = 0$ in the context of Newtonian mechanics can either represent a body in an inertial frame under linear drag or a body in a non-inertial frame with no forces acting on it. If $t_{in}$ is time in the inertial frame, the time $t$ in the non-inertial frame is such that  $\frac{dt}{dt_{in}} = e^{\frac{b}{m}t}$. Mathematical equivalence does not imply physical equivalence, which is especially problematic for~\cite{Barrett2} since it focuses on the former; for similar arguments see~\cite{Coffey}. In other words, we can have the same physics with different math and the same math with different physics. The connection between math and physics is many-to-many because mathematics captures only the syntax of a particular description without its semantics, which at that point cannot be recovered.

As we do not have a one-to-one map between math and physics, fixating on the mathematical structure may even confuse us. For example, it is known~\cite{AllHamFreeParticle} that any Hamiltonian system is, at least locally, equivalent to a free particle. That is, we can always find a canonical transformation (i.e. a transformation that does not change the equation of motion) such that any Hamiltonian system, in the new coordinates, has the motion of a free particle for a finite amount of time. As~\cite{North} notes, only frame-independent relationships should be taken as fundamental. Does that mean that the Hamiltonian itself is not fundamental because it is not invariant under canonical transformations? Does it mean that all Hamiltonian systems are the same system, since there is always a set of canonical coordinates in which $H=p^2/2m$? It is also known~\cite{AllSystemsAreHam} that for any first-order system of ordinary differential equations we can find an appropriate symplectic structure such that the system can be written as a Hamiltonian system. Does that mean that, in the end, all systems are physically equivalent, regardless of what they describe? The point is, we cannot know whether our findings about mathematical structures are physically significant if we do not have a precise link between the mathematical symbols and their physical meaning.

This leads us to the second problem we believe the previous works may have: when trying to attach meaning the thinking and intuition is still Newtonian. This is particularly evident in~\cite{Curiel} where physical ideas provide the starting point to get to Lagrangian dynamics. The starting point is that a classical system is one that, roughly, obeys Newton's second law and therefore we have, for a free particle, $\dot{x} = v$ and $\dot{v} = 0$. Unfortunately this is only true in an inertial frame. As~\cite{North} notes, both Hamiltonian and Lagrangian mechanics are coordinate independent: the equations are valid in all reference frames even non-inertial ones. In fact, that is probably the most notable advantage of the frameworks, that fictitious forces are automatically handled. But none of the Newtonian concepts are coordinate independent: inertial frames, absence of forces and even conservation of energy are all frame dependent. This, again, leads to confusion and we may miss a key element: in the transition from Newtonian to Lagrangian/Hamiltonian mechanics we expand the reference frames in which the equations are valid (from inertial frames to all frames) at the expense of restricting the class of systems we can describe (e.g. dissipative forces in general require a modification of the Euler-Lagrange equation typically using the Rayleigh dissipation function). The point is, Newtonian concepts may prevent us from fully understanding Hamiltonian and Lagrangian mechanics.

It should be clear that we do not fault the authors. Even just a century ago, there was a push in physics to find fundamental principles or laws that could serve as a starting point for the different theories. Newtonian mechanics, thermodynamics and special relativity serve as good examples. Nowadays, unfortunately, many of the current fundamental theories, including Lagrangian, Hamiltonian, quantum mechanics and quantum field theory take as their starting point a particular mathematical structure. That is, we do not know what the value of a Lagrangian means or what the symplectic form represents; we just take them as given. Moreover, a lot of work in theoretical physics nowadays starts by exploring novel abstract mathematical structures and then trying to see whether some physical meaning or prediction can be extracted. We are not going to debate here how and why this shift happened. We merely note that it leaves the semantics of our theories ill defined.\footnote{Which is a nice way to say that we don't really know what we are talking about.}

@@

As all these problems stem from a lack of physical principles that can serve as a foundation for classical Hamiltonian and Lagrangian mechanics, we turn to a recently published work~\cite{AoPPhy1} that aims to identify such starting points.\footnote{We should also mention~\cite{Maudlin} as another attempt to formally derive mathematical structure from physical ideas.} Inspired by those insights, we fully develop another that is only briefly hinted at in that work: the idea that \textbf{a distribution over a manifold conserves information entropy over time if and only if its elements evolve according to Hamiltonian mechanics.} That is, the single requirement of information entropy conservation not only gives us Hamiltonian dynamics but also the structure of phase space (i.e. a symplectic structure where variables are organized in conjugate pairs).

This insight helps to clearly characterize what Hamiltonian mechanics describes, providing a clear physical meaning to the mathematical structures used. This should be at least of interest to that part of the philosophy community interested in the foundations of classical mechanics. Second because information entropy has a deep connection to the foundations of thermodynamics, therefore the strong link we establish here between information entropy and Hamiltonian mechanics should be of interest to an even wider audience.

@@

Before proceeding, though, let us give a quick introduction so that we have a clear understanding about what information entropy represents. In the physics community, for example, the concept is often vaguely referred to as ``knowledge" (and sometimes as ``lack of knowledge") which is incorrect and misleading.\footnote{We believe this characterization may be due to Jaynes~\cite{Jaynes} who first introduced the concept of information theory within thermodynamics. Note that we are only going to talk about information entropy and not thermodynamic entropy. We do believe there is a link between the two, but it is not the one Jaynes gives, which we believe to be unphysical. We leave that discussion for later works.} Information entropy was introduced by Shannon~\cite{Shannon} to solve specific engineering problems and therefore, like distance or mass, information entropy is a number that measures a well defined quantity and should be treated as such. What he needed was a way to measure how much information a particular channel is able to transfer and therefore a way to quantify precisely information as number of bits. So the idea is that we have a source that sends messages to a destination, and the message is encoded in zeros and ones. Information entropy is the optimal average number of zeros and ones that must be used to encode the messages. If the source sends fewer bits per message than required by the information entropy, some information will be lost. If the source sends more bits per message than required by the information entropy, some information will be redundant.

@@

In general, if $I$ is the number of questions, the number of bits, then $C = 2 ^ I$ are the number of cases one can distinguish with those questions. On the other hand, suppose for a particular value $\xi_0$ of a discrete variable we have $\rho(\xi_0)=1/5$, then we know that we'd select $\xi_0$ one in 5 cases. That is, the normalized distribution is the inverse of the number of cases: $\rho(\xi_0) = \frac{1}{C}$.  Therefore the information needed to identify a particular case is $I(\xi^a)=\log \frac{1}{\rho(\xi^a)}$. If we take the expectation value of this expression over the distribution and change the base of the logarithm, we have:
\begin{equation}
I(\rho(\xi^a)) = \int \rho(\xi^a) \ln \left(\frac{1}{\rho(\xi^a)}\right) d\xi^1 d\xi^2 ... d\xi^n =-\int \rho(\xi^a) \ln (\rho(\xi^a)) d\xi^1 d\xi^2 ... d\xi^n
\end{equation}
which is the information entropy of the distribution. This represents the average number of bits one needs to put into a message to tell someone else the value of the element of the distribution that was picked at random.

@@

The above discussion, while not a formal derivation, should give a sense of what the expression signifies: it measures the \emph{additional} information that would be needed to identify the value associated to an element picked at random from a known population. Distributions that are more spread will have a higher information entropy. What is also important is that it is a precisely defined quantity and that there is only one way to define it. It is the only expression that satisfies three simple requirements (continuity, monotonicity and additivity) that are necessary to measure information, and that is how Shannon originally derived it. Insofar that we want to talk about amounts of information we have only one way to do so.

@@

We now turn our attention to what~\cite{AoPPhy1} calls an infinitesimally reducible system. This is a system that can be repeatedly divided into smaller and smaller parts and for which giving a description of the full system is equivalent to giving a full description for all parts. We call particles the limit of this recursive subdivision. Note that particles here are not point-like: they have an infinitesimal size for each dimension along which the system was subdivided. Under this premise each particle is assigned a state from a state space $\mathcal{S}$ and the state of the whole system will be described by a distribution over those particle states. The elements of our distribution are the particles, and the property that characterizes each element is its state.

@@

While the physical setting is clear enough, it introduces a puzzle. States will be themselves identified by quantities, which we call state variables. But for states to be physically objective, they must be independent of reference frame. For example, if $s$ is the state, $A$ and $B$ are two reference frames and $\xi^a$ and $\hat{\xi}^b$ are the state variables one would use in the respective reference frames, we must have $s(\xi^a) = s(\hat{\xi}^b)$.  Therefore, since the distribution is defined on the state itself, we must have $\rho(s(\xi^a))=\rho(s(\hat{\xi}^b))$. In this case, it seems we \emph{do} want the density to be defined on the actual space $\mathcal{S}$, and it seems it will have to be invariant under an arbitrary change of state variables. Accordingly, if we want densities and information entropy to be conserved over time, we first need to make sure these quantities are defined in a way that is observer independent, since otherwise they would be conserved in one frame but not in another. But this seems in apparent contradiction with the previous section and we need to clarify that.

@@

If we want the time evolution to transport  densities and to conserve information entropy, the Jacobian must be unitary and therefore $S$ is divergence-free and admits a potential. We have
\begin{equation}
\label{Potential_Hamilton}
\begin{aligned}
S &= \left(\frac{\partial H}{\partial p}, - \frac{\partial H}{\partial q}\right) \\
\frac{dq}{dt} &= \frac{\partial H}{\partial p}  \\
\frac{dp}{dt} &= - \frac{\partial H}{\partial q}  \\
\end{aligned}
\end{equation}
and we recognize Hamilton's equations. Note that the reverse is true as well: any Hamiltonian evolution will transport densities and conserve information entropy due to Liouville's theorem.\footnote{Again, it is not just that the density and the areas are conserved in time: it is that they are the same for all canonical transformations including change of coordinates. That is, we have an objective way to compare them at different points.} Therefore \textbf{Hamiltonian mechanics is precisely the evolution of (classical) distributions over a manifold that preserves information entropy and densities}.

@@

%Now, the fact that we can use a tangent bundle to track our particle states does \emph{not} mean that conjugate momentum is a one form, in the sense of a map from a vector to a number. The state space is just a manifold, a set of points identified by $n$ continuous variable, and the structure it has comes from how units transform, not because the variables identify separate objects. Unfortunately mathematical structures strip out the units, the knowledge of the double role of coordinates is lost and then a different mathematical concept is used, one that has the same mathematical features of our particle state space but an altogether different meaning and construction. Changing state variables from position and momentum to position and velocity is not, as one may mistakenly understand from the math, a change of objects from a cotangent bundle to a tangent bundle: it is relabeling the same states much like changing spatial coordinates relabels positions. This change of variables, though, is not in general a canonical transformation which means the density will not correspond to the invariant one. But it is still a valid change of state variables and in fact, giving that velocity is gauge invariant, a very useful one. It allows to write Poisson brackets and find invariants of motion that are readily physically meaningful~\cite{Holten}.

@@

Now, the fact that we can use symplectic manifolds, or tangent bundles, to track our particle states does \emph{not} mean that conjugate momentum is a one form, in the sense of a map from a vector to a number, defined on the coordinate space. The state space comes as one whole manifold, a set of points identified by $n$ continuous variables $\xi^a$, and the structure it has comes from how units transform, not because the variables identify separate objects. Unfortunately, mathematical structures strip out the units, the knowledge of the double role of coordinates is lost and then a different mathematical concept is used, one that has the same mathematical features of our particle state space but an altogether different meaning and construction. We could, in many cases, use position and velocity as state variables. While these would not be canonical coordinates and thus not a good choice to track our densities, they are still valid state variables and in fact, given that velocity is gauge invariant, very useful ones~\cite{Holten}. But the space has not changed: we still have states. The tangent and cotangent bundle structures, then, are just tracking how conjugate momentum and velocity change differently under coordinate transformation.

@@

This approach clarifies what we believe is a source of confusion. One may think that conceptually starting with parts and aggregating them to a whole is equivalent to starting from a whole and dividing it into parts. This is true only in the finite case. In the infinite case we have to start with what is finite (either the whole or the parts) and take the limit (either division or aggregation). We contend that the top-down approach (i.e. we start with a finite whole and keep subdividing) is the correct one to understand classical mechanics.

First because it maps better to what we do in classical physics. We study composite systems (planets, cannonballs, beads on a wire, fluids) which we pretend we can recursively divide. None of those objects has a single position or momentum: they have an (idealized) center of mass with position and momentum given by the expectation values $\int q \rho(q,p) dq dp$ and $\int p \rho(q,p) dq dp$ of the distribution of its parts. A ``point particle'' is really a peaked distribution whose extent can be disregarded for the problem at hand.

Second because it solves a set of conceptual problems: if we start with point-like objects of zero size, how can we get a finite size? Shouldn't we have a discrete topology? How could we have a differentiable structure? We contend that measure, topologies, metrics, differentiable structure and so on are there to remind us that the points are infinitesimal limits, keep track of how they were taken and which aggregations of those limits correspond to a finite whole.

Third because it is conceptually more compatible with the other physical theories. We have seen how elements of special relativity (e.g. coordinate invariance, momentum as covector), statistical mechanics (e.g. joint and marginal distributions), quantum mechanics (e.g. wave number) and classical Hamiltonian mechanics (e.g. phase space, Poisson brackets) already emerged naturally. Furthermore, note that in general relativity point-like particles would be singularities that puncture space-time and would not move along geodesics, while particles as infinitesimal parts would. Note that if the evolution is deterministic and reversible, it means the evolution of the system depends only on itself, which means that the system is isolated. That is, we have an association between isolated system and conservation of energy.\footnote{To be precise, conservation of energy is not an invariant property. The fact that a Hamiltonian exists, though, is an invariant property.} This sounds strikingly close to the first law of thermodynamics.

The larger point, then, is that physics is one and it cannot truly be understood piecemeal. Nature does not separate between dynamical systems and thermodynamics, between topologies and $\sigma$-algebras, between classical systems and quantum systems. The boundaries between all these disciplines are as much unfortunate as they are artificial. There is no real understanding of one without the other. This approach helps to resolve this by giving us a more coherent picture that spans across the different areas, which is not only intellectually more satisfying but it allows us to switch perspective when the intuition from one fails. As a practical example, we can show how understanding the link between Hamiltonian mechanics and information theory leads to a classical analogue for the uncertainty principle.

@@

As we saw, Hamiltonian dynamics not only conserves the energy of individual particles, but it also conserves the information entropy of the distribution as a whole. Intuitively, we can see that there is some link between the spread of a distribution and the number of bits required to identify an element. For example, if we have two uniform distributions with different ranges, then Alice will need to give Bob more information to identify the value from the bigger range. So we ask: what is the distribution that minimizes the spread given a certain amount of information entropy? If $I$ is the set amount of information entropy and $\sigma_q^2 \sigma_p^2 \equiv \int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp$ is the spread, we can use Lagrange multipliers to find out.
\begin{align*}
L = &\int (q-\mu_q)^2 \rho \, dqdp \int (p-\mu_p)^2 \rho \, dqdp \\
&+ \lambda_1(\int \rho dqdp - 1) \\ &+ \lambda_2(- \int \rho \ln \rho \, dqdp - I)\\ 
\delta L = &\int \delta \rho [(q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 + \\ &\lambda_1 - \lambda_2 \ln \rho - \lambda_2 ] dqdp = 0 \\
\lambda_2 \ln \rho = &\lambda_1 - \lambda_2 + (q-\mu_q)^2 \sigma_p^2 + \sigma_q^2 (p-\mu_p)^2 \\
\rho = &e^{\frac{\lambda_1 - \lambda_2}{\lambda_2}}e^{\frac{(q-\mu_q)^2 \sigma_p^2}{\lambda_2}}e^{\frac{\sigma_q^2 (p-\mu_p)^2}{\lambda_2}}\\
\end{align*}
We solve the multipliers and have:
\begin{align*}
\rho = &\frac{1}{ 2 \pi \sigma_q \sigma_p} e^{-\frac{(q-\mu_q)^2}{2\sigma_q^2}} e^{-\frac{(p-\mu_p)^2}{2\sigma_p^2}} \\
I = &\ln (2\pi\sigma_q\sigma_p) + 1
\end{align*}
The distribution that minimizes the spread, then, is the product of two independent Gaussians. As the entropy is conserved during Hamiltonian evolution the product $\sigma_q^2 \sigma_p^2$ can never be less than the one given by the Gaussian distribution of the same entropy. We have:
\begin{align*}
\sigma_q\sigma_p \geq \exp (I - 1) / 2 \pi 
\end{align*}
Hamiltonian mechanics is the deterministic and reversible evolution for a distribution and its elements, and therefore it cannot reduce the spread of the distribution indefinitely. If it did, it would start concentrating the density, meaning that multiple states would be mapped to a single state and the dynamics would not be reversible. In this form $I$ is not a fundamental constant but a parameter that depends upon the initial condition. While this is already striking,~\cite{AoPPhy1} shows a simple argument to turn this into something more.

Suppose that we have a distribution where the particles move non-deterministically. That is, their motion is partly caused by some other agent in a way that can be characterized by random motion. Now suppose our distribution is at an equilibrium with that random motion. Then the particles change state randomly but the overall distribution remains the same. That is, the random motion simply reshuffles the elements. In this case the minimum achievable for the information entropy of the distribution depends on the random motion itself. The stronger it is, the more spread the distribution has to be, the higher the minimum threshold for the information entropy.
